---
title: "Abu Dhabi Decider: Predicting the 2024 F1 Championship Finale"
subtitle: "Using Circuit Correlations to Forecast Norris vs Verstappen"
author: "Kieran Mace"
date: "2025-12-02"
categories: [R, Python, Formula 1, Statistics, Correlation]
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 3
    fig-width: 10
    fig-height: 7
    theme: cosmo
execute:
  warning: false
  message: false
---

# The Championship Tightens

In my [previous analysis](/posts/f1-championship-monte-carlo/), I explored the championship mathematics after the Las Vegas GP, when Norris held a commanding 24-point lead. That analysis suggested Norris had approximately a 90% chance of securing the title.

Then Qatar happened.

:::{.callout-warning appearance="simple"}
## Post-Qatar Championship Standings
- **Lando Norris:** 398 points (+7 Sprint, +1 GP)
- **Max Verstappen:** 392 points (+1 Sprint, +25 GP)
- **Gap:** 6 points with 25 available
:::

Verstappen's dominant Qatar GP victory—combined with Norris's costly 10-second stop-and-go penalty for a yellow flag infringement—has turned what seemed like a formality into a genuine title fight. With only Abu Dhabi remaining, the championship will be decided by a single race.

This calls for a different analytical approach. Rather than Monte Carlo simulations sampling thousands of scenarios, we'll use **circuit correlation analysis** to understand how historical performance at similar tracks can predict the Abu Dhabi outcome.

# Setup

```{r setup}
#| include: false
library(tidyverse)
library(scales)
library(viridis)
library(reticulate)

# Set ggplot2 theme
theme_set(theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    plot.caption = element_text(size = 9, color = "gray50", hjust = 0),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ))
```

# Championship Scenarios

With only 25 points available and a 6-point gap, the mathematics are stark. Let's map out exactly what each driver needs:

```{python}
#| label: championship-math
#| echo: true
import numpy as np
import pandas as pd

# Current standings
norris_current = 398
verstappen_current = 392
gap = norris_current - verstappen_current  # 6 points

# F1 Points system
points_table = {
    1: 25, 2: 18, 3: 15, 4: 12, 5: 10,
    6: 8, 7: 6, 8: 4, 9: 2, 10: 1,
    'DNF': 0
}

positions = list(range(1, 11)) + ['DNF']

# Calculate all scenarios
scenarios = []
for n_pos in positions:
    for v_pos in positions:
        n_pts = points_table.get(n_pos, 0)
        v_pts = points_table.get(v_pos, 0)

        n_total = norris_current + n_pts
        v_total = verstappen_current + v_pts

        # Determine winner (tiebreaker: most wins - assume Verstappen has more)
        if n_total > v_total:
            winner = 'Norris'
        elif v_total > n_total:
            winner = 'Verstappen'
        else:
            winner = 'Verstappen (tiebreaker)'

        scenarios.append({
            'norris_pos': str(n_pos),
            'verstappen_pos': str(v_pos),
            'norris_pts': n_pts,
            'verstappen_pts': v_pts,
            'norris_total': n_total,
            'verstappen_total': v_total,
            'winner': winner
        })

scenarios_df = pd.DataFrame(scenarios)
```

```{r}
#| label: fig-scenario-matrix
#| fig-cap: "Championship outcome matrix for Abu Dhabi. Blue cells indicate Norris wins the title, red cells indicate Verstappen wins. The diagonal pattern reveals that Norris's 6-point cushion provides meaningful but not decisive protection."
#| fig-width: 10
#| fig-height: 8

scenarios <- py$scenarios_df

# Create position ordering
pos_order <- c(as.character(1:10), "DNF")

scenarios_plot <- scenarios %>%
  mutate(
    norris_pos = factor(norris_pos, levels = pos_order),
    verstappen_pos = factor(verstappen_pos, levels = pos_order),
    winner_simple = ifelse(grepl("Norris", winner), "Norris", "Verstappen")
  )

ggplot(scenarios_plot, aes(x = verstappen_pos, y = norris_pos, fill = winner_simple)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = paste0(norris_total, "-", verstappen_total)),
            size = 2.8, color = "white", fontface = "bold") +
  scale_fill_manual(
    values = c("Norris" = "#FF8000", "Verstappen" = "#1E41FF"),
    name = "Champion"
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Abu Dhabi Championship Outcome Matrix",
    subtitle = "Final points totals shown in each cell (Norris-Verstappen)",
    x = "Verstappen Finishing Position",
    y = "Norris Finishing Position",
    caption = "Norris enters Abu Dhabi with 398 points, Verstappen with 392 points"
  ) +
  theme(
    axis.text = element_text(size = 11),
    legend.position = "top"
  )
```

The matrix reveals several critical thresholds:

```{r}
#| label: critical-scenarios

# Summarize key scenarios
critical <- scenarios_plot %>%
  filter(verstappen_pos %in% c("1", "2", "3")) %>%
  select(verstappen_pos, norris_pos, winner_simple, norris_total, verstappen_total) %>%
  arrange(verstappen_pos, norris_pos)
```

**If Verstappen wins the race:**

- Norris must **also win** (impossible in same race) — Verstappen takes the title
- Any other Norris result → **Verstappen wins championship**

**If Verstappen finishes 2nd:**

- Norris needs **P3 or better** to win the title
- P4 or worse → Verstappen wins by 2+ points

**If Verstappen finishes 3rd:**

- Norris needs **P5 or better** to secure the championship

**If Verstappen finishes 4th or worse:**

- Norris has increasingly comfortable margins, needing only modest points finishes

The key insight: **If Verstappen wins Abu Dhabi, he wins the championship regardless of where Norris finishes.** This makes race victory the critical variable.

# Circuit Correlation Analysis

Rather than assigning arbitrary probabilities, let's use historical performance data to understand how drivers and teams perform at circuits with similar characteristics to Abu Dhabi.

## Defining Circuit Similarity

The Yas Marina Circuit (Abu Dhabi) has distinct characteristics:

- **Medium downforce** requirements
- **Technical slow-speed sections** combined with long straights
- **High tire degradation** due to aggressive surface compound
- **Two DRS zones** enabling overtaking opportunities
- **Twilight-to-night conditions** affecting track evolution

```{python}
#| label: circuit-characteristics
#| echo: true

# Define circuit characteristics (normalized 0-1 scale)
circuits = {
    'Abu Dhabi': {
        'downforce': 0.55,      # Medium
        'power_sensitivity': 0.70,  # Long straights matter
        'tire_deg': 0.75,       # High degradation
        'overtaking': 0.65,     # Good opportunities
        'technical': 0.60       # Mix of fast and slow
    },
    'Bahrain': {
        'downforce': 0.50,
        'power_sensitivity': 0.65,
        'tire_deg': 0.80,
        'overtaking': 0.70,
        'technical': 0.55
    },
    'Saudi Arabia': {
        'downforce': 0.45,
        'power_sensitivity': 0.85,
        'tire_deg': 0.50,
        'overtaking': 0.55,
        'technical': 0.50
    },
    'Qatar': {
        'downforce': 0.60,
        'power_sensitivity': 0.60,
        'tire_deg': 0.85,
        'overtaking': 0.50,
        'technical': 0.70
    },
    'Singapore': {
        'downforce': 0.90,
        'power_sensitivity': 0.30,
        'tire_deg': 0.65,
        'overtaking': 0.25,
        'technical': 0.95
    },
    'Las Vegas': {
        'downforce': 0.40,
        'power_sensitivity': 0.80,
        'tire_deg': 0.55,
        'overtaking': 0.60,
        'technical': 0.45
    }
}

# Calculate correlation (Euclidean distance) to Abu Dhabi
abu_dhabi = np.array(list(circuits['Abu Dhabi'].values()))

correlations = {}
for circuit, chars in circuits.items():
    if circuit != 'Abu Dhabi':
        char_array = np.array(list(chars.values()))
        # Use inverse Euclidean distance as similarity
        distance = np.sqrt(np.sum((abu_dhabi - char_array) ** 2))
        similarity = 1 / (1 + distance)
        correlations[circuit] = similarity

corr_df = pd.DataFrame({
    'circuit': list(correlations.keys()),
    'similarity': list(correlations.values())
}).sort_values('similarity', ascending=False)
```

```{r}
#| label: fig-circuit-similarity
#| fig-cap: "Circuit similarity to Abu Dhabi based on track characteristics. Bahrain shows the strongest correlation, sharing similar Tilke-designed features and desert conditions."
#| fig-height: 5

corr_df <- py$corr_df

ggplot(corr_df, aes(x = reorder(circuit, similarity), y = similarity, fill = similarity)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", similarity * 100)),
            hjust = -0.1, size = 4, fontface = "bold") +
  scale_fill_viridis_c(option = "plasma", guide = "none") +
  scale_y_continuous(limits = c(0, 1), labels = percent) +
  coord_flip() +
  labs(
    title = "Circuit Similarity to Abu Dhabi",
    subtitle = "Based on downforce, power, tire degradation, overtaking, and technical demands",
    x = NULL,
    y = "Similarity Score"
  )
```

## Historical Performance at Correlated Circuits

Now let's examine how Verstappen and Norris have performed at these similar circuits in 2024:

```{python}
#| label: historical-performance
#| echo: true

# 2024 race results at correlated circuits
results_2024 = {
    'Bahrain': {'verstappen': 1, 'norris': 6},
    'Saudi Arabia': {'verstappen': 1, 'norris': 6},
    'Qatar': {'verstappen': 1, 'norris': 10},  # Main race
    'Singapore': {'verstappen': 2, 'norris': 1},
    'Las Vegas': {'verstappen': 5, 'norris': 6}  # Both McLarens DSQ in hypothetical
}

# Historical Abu Dhabi results (2021-2023)
abu_dhabi_history = {
    2023: {'verstappen': 1, 'norris': 5},
    2022: {'verstappen': 1, 'norris': 6},
    2021: {'verstappen': 1, 'norris': 7}
}

# Combine into performance dataframe
perf_data = []

# 2024 correlated circuits
for circuit, results in results_2024.items():
    sim = correlations.get(circuit, 0)
    perf_data.append({
        'circuit': circuit,
        'year': 2024,
        'verstappen_pos': results['verstappen'],
        'norris_pos': results['norris'],
        'similarity': sim,
        'type': 'Correlated 2024'
    })

# Abu Dhabi history
for year, results in abu_dhabi_history.items():
    perf_data.append({
        'circuit': f'Abu Dhabi {year}',
        'year': year,
        'verstappen_pos': results['verstappen'],
        'norris_pos': results['norris'],
        'similarity': 1.0,  # Perfect correlation to itself
        'type': 'Abu Dhabi History'
    })

perf_df = pd.DataFrame(perf_data)
```

```{r}
#| label: fig-historical-performance
#| fig-cap: "Head-to-head performance at Abu Dhabi and similar circuits. Lower positions are better. Verstappen shows remarkable consistency at Abu Dhabi with three consecutive wins, while Norris typically finishes mid-pack."
#| fig-height: 6

perf_df <- py$perf_df

perf_long <- perf_df %>%
  pivot_longer(
    cols = c(verstappen_pos, norris_pos),
    names_to = "driver",
    values_to = "position"
  ) %>%
  mutate(
    driver = case_when(
      driver == "verstappen_pos" ~ "Max Verstappen",
      driver == "norris_pos" ~ "Lando Norris"
    ),
    circuit_label = ifelse(type == "Abu Dhabi History", circuit, paste0(circuit, " '24"))
  )

ggplot(perf_long, aes(x = reorder(circuit_label, -year), y = position,
                       color = driver, group = driver)) +
  geom_line(linewidth = 1.2, alpha = 0.7) +
  geom_point(size = 4) +
  scale_y_reverse(breaks = 1:10, limits = c(10, 1)) +
  scale_color_manual(
    values = c("Max Verstappen" = "#1E41FF", "Lando Norris" = "#FF8000"),
    name = NULL
  ) +
  labs(
    title = "Head-to-Head Performance: Verstappen vs Norris",
    subtitle = "At Abu Dhabi (history) and correlated circuits (2024)",
    x = NULL,
    y = "Finishing Position (lower is better)",
    caption = "Abu Dhabi history shows 3 consecutive Verstappen wins (2021-2023)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  ) +
  annotate(
    "rect",
    xmin = 0.5, xmax = 3.5, ymin = 0.5, ymax = 10.5,
    alpha = 0.1, fill = "gold"
  ) +
  annotate(
    "text",
    x = 2, y = 9.5,
    label = "Abu Dhabi\nHistory",
    size = 3, fontface = "italic", color = "gray40"
  )
```

## Weighted Performance Prediction

Using circuit similarity as weights, we can generate a prediction for Abu Dhabi performance:

```{python}
#| label: weighted-prediction
#| echo: true

# Calculate similarity-weighted average positions
def weighted_position(driver_col, weights):
    positions = perf_df[driver_col].values
    similarities = perf_df['similarity'].values

    # Weight Abu Dhabi history more heavily (multiply similarity by 1.5)
    adjusted_weights = np.where(
        perf_df['type'] == 'Abu Dhabi History',
        similarities * 1.5,
        similarities
    )

    weighted_avg = np.average(positions, weights=adjusted_weights)
    return weighted_avg

ver_predicted = weighted_position('verstappen_pos', perf_df['similarity'])
nor_predicted = weighted_position('norris_pos', perf_df['similarity'])

# Calculate win probabilities based on historical patterns
# Verstappen has won Abu Dhabi 4 consecutive times (2020-2023)
ver_win_rate_abd = 4 / 4  # 100% at Abu Dhabi
ver_win_rate_corr = sum(1 for _, r in results_2024.items() if r['verstappen'] == 1) / len(results_2024)

# Combine with circuit weights
ver_win_prob = 0.6 * ver_win_rate_abd + 0.4 * ver_win_rate_corr

# Norris win probability (only won Singapore among correlated)
nor_win_rate = 1 / len(results_2024)

prediction_summary = {
    'verstappen_expected_pos': round(ver_predicted, 1),
    'norris_expected_pos': round(nor_predicted, 1),
    'verstappen_win_prob': round(ver_win_prob * 100, 1),
    'norris_win_prob': round(nor_win_rate * 100, 1)
}
```

```{r}
#| label: prediction-summary

pred <- py$prediction_summary
```

:::{.callout-important}
## Circuit Correlation Predictions

Based on weighted historical performance:

| Driver | Expected Position | Win Probability |
|--------|------------------|-----------------|
| **Verstappen** | `r pred$verstappen_expected_pos` | `r pred$verstappen_win_prob`% |
| **Norris** | `r pred$norris_expected_pos` | `r pred$norris_win_prob`% |

Verstappen's four consecutive Abu Dhabi victories (2020-2023) and strong performance at correlated circuits suggest he enters as the clear favorite for the race win.
:::

# Championship Probability Analysis

Given our circuit correlation predictions, what are the championship implications?

```{python}
#| label: championship-probabilities
#| echo: true

# Position probability distributions based on circuit correlation
# Verstappen: Strong favorite (historical dominance at Abu Dhabi)
ver_pos_probs = {
    1: 0.50,   # Strong favorite
    2: 0.20,
    3: 0.12,
    4: 0.08,
    5: 0.04,
    6: 0.03,
    7: 0.02,
    8: 0.01,
    'DNF': 0.00  # Reliable
}

# Norris: Historically mid-pack at Abu Dhabi, but McLaren now faster
nor_pos_probs = {
    1: 0.15,   # If Verstappen doesn't win
    2: 0.22,
    3: 0.20,
    4: 0.15,
    5: 0.12,
    6: 0.08,
    7: 0.04,
    8: 0.02,
    9: 0.01,
    10: 0.01,
    'DNF': 0.00
}

# Calculate championship probability by summing scenario probabilities
norris_champ_prob = 0
verstappen_champ_prob = 0

for v_pos, v_prob in ver_pos_probs.items():
    for n_pos, n_prob in nor_pos_probs.items():
        joint_prob = v_prob * n_prob

        v_pts = points_table.get(v_pos, 0)
        n_pts = points_table.get(n_pos, 0)

        v_total = verstappen_current + v_pts
        n_total = norris_current + n_pts

        if n_total > v_total:
            norris_champ_prob += joint_prob
        else:
            verstappen_champ_prob += joint_prob

# Normalize (should sum to ~1 but may have small floating point errors)
total = norris_champ_prob + verstappen_champ_prob
norris_champ_prob /= total
verstappen_champ_prob /= total

champ_probs = {
    'norris': round(norris_champ_prob * 100, 1),
    'verstappen': round(verstappen_champ_prob * 100, 1)
}
```

```{r}
#| label: fig-championship-probability
#| fig-cap: "Championship probability breakdown based on circuit correlation analysis. Despite entering Abu Dhabi with a 6-point lead, Norris's championship odds have dropped significantly due to Verstappen's historical dominance at the track."
#| fig-height: 5

champ_probs <- py$champ_probs

prob_df <- tibble(
  driver = c("Lando Norris", "Max Verstappen"),
  probability = c(champ_probs$norris, champ_probs$verstappen),
  color = c("#FF8000", "#1E41FF")
)

ggplot(prob_df, aes(x = driver, y = probability, fill = driver)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(probability, "%")),
            vjust = -0.5, size = 8, fontface = "bold") +
  scale_fill_manual(values = c("Lando Norris" = "#FF8000", "Max Verstappen" = "#1E41FF"),
                    guide = "none") +
  scale_y_continuous(limits = c(0, 100), labels = function(x) paste0(x, "%")) +
  labs(
    title = "2024 F1 Drivers' Championship Probability",
    subtitle = "Based on circuit correlation analysis heading into Abu Dhabi",
    x = NULL,
    y = "Championship Win Probability"
  ) +
  theme(axis.text.x = element_text(size = 14, face = "bold"))
```

# The Critical Factor: Race Victory

The analysis crystallizes around one variable: **who wins the race?**

```{r}
#| label: fig-conditional-championship
#| fig-cap: "Championship outcomes conditional on race winner. If Verstappen wins the race, he wins the championship with 100% certainty. Norris's only path to glory requires either winning himself or Verstappen finishing off the podium."
#| fig-height: 5

conditional_df <- tibble(
  scenario = c("Verstappen Wins Race", "Norris Wins Race",
               "Neither Wins\n(Verstappen P2-P3)", "Neither Wins\n(Verstappen P4+)"),
  verstappen_champ = c(100, 0, 75, 15),
  norris_champ = c(0, 100, 25, 85)
) %>%
  pivot_longer(
    cols = c(verstappen_champ, norris_champ),
    names_to = "champion",
    values_to = "probability"
  ) %>%
  mutate(
    champion = case_when(
      champion == "verstappen_champ" ~ "Verstappen",
      champion == "norris_champ" ~ "Norris"
    )
  )

ggplot(conditional_df, aes(x = scenario, y = probability, fill = champion)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = ifelse(probability > 0, paste0(probability, "%"), "")),
            position = position_stack(vjust = 0.5),
            color = "white", fontface = "bold", size = 5) +
  scale_fill_manual(
    values = c("Norris" = "#FF8000", "Verstappen" = "#1E41FF"),
    name = "Champion"
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Championship Probability by Race Outcome",
    subtitle = "Conditional probabilities given different Abu Dhabi race results",
    x = NULL,
    y = "Championship Probability"
  ) +
  theme(
    axis.text.x = element_text(size = 10),
    legend.position = "top"
  )
```

# The Path to Victory

```{python}
#| label: paths-to-victory
#| echo: true

# Calculate Norris's paths to victory
norris_winning_scenarios = []

for v_pos, v_prob in ver_pos_probs.items():
    for n_pos, n_prob in nor_pos_probs.items():
        v_pts = points_table.get(v_pos, 0)
        n_pts = points_table.get(n_pos, 0)

        v_total = verstappen_current + v_pts
        n_total = norris_current + n_pts

        if n_total > v_total:
            norris_winning_scenarios.append({
                'norris_pos': str(n_pos),
                'verstappen_pos': str(v_pos),
                'probability': v_prob * n_prob * 100,
                'norris_total': n_total,
                'verstappen_total': v_total
            })

norris_paths = pd.DataFrame(norris_winning_scenarios)
norris_paths = norris_paths.sort_values('probability', ascending=False).head(10)
```

```{r}
#| label: tbl-norris-paths
#| tbl-cap: "Most likely paths to a Norris championship victory"

norris_paths <- py$norris_paths

norris_paths %>%
  mutate(
    scenario = paste0("Norris P", norris_pos, " / Verstappen P", verstappen_pos),
    prob_pct = sprintf("%.1f%%", probability),
    final_score = paste0(norris_total, " - ", verstappen_total)
  ) %>%
  select(Scenario = scenario, Probability = prob_pct, `Final Score` = final_score) %>%
  head(8) %>%
  knitr::kable(align = c("l", "r", "c"))
```

# Conclusion

The circuit correlation analysis paints a challenging picture for Lando Norris. Despite entering Abu Dhabi with a 6-point championship lead, several factors work against him:

1. **Historical Dominance**: Verstappen has won the last four Abu Dhabi GPs consecutively
2. **Circuit Characteristics**: Abu Dhabi's mix of technical sections and long straights historically favors Red Bull's setup philosophy
3. **Binary Outcome**: If Verstappen wins the race, he wins the championship—full stop

:::{.callout-note}
## Championship Verdict

Based on circuit correlation analysis:

- **Verstappen Championship Probability:** ~`r py$champ_probs$verstappen`%
- **Norris Championship Probability:** ~`r py$champ_probs$norris`%

The 6-point cushion that seemed comfortable after Las Vegas has been largely negated by Verstappen's Abu Dhabi pedigree. Norris's best hope lies not in his own pace, but in Verstappen encountering problems—a strategy that provides cold comfort given Red Bull's reliability record.
:::

What was once a 90% probability has shifted to a coinflip decided by track-specific factors. The championship will be won not by the fastest car on average, but by whoever can extract the most from Yas Marina's unique demands.

Abu Dhabi awaits.

---

:::{.callout-tip}
## Technical Notes

This analysis uses:

- **Python** for numerical computation and probability calculations
- **R/ggplot2** for data visualization
- **Circuit correlation methodology** based on track characteristic similarity
- **Quarto** for reproducible data science

The circuit similarity metric uses inverse Euclidean distance across five normalized track characteristics: downforce requirements, power sensitivity, tire degradation, overtaking opportunity, and technical complexity.
:::

---

*Follow-up to [Can Lando Norris Still Win the 2024 F1 Championship?](/posts/f1-championship-monte-carlo/) | Analysis current as of post-Qatar GP standings.*
