---
title: "Illinois Firearm Safe Storage Survey Analysis 2023"
subtitle: "Understanding Awareness, Practices, and Demographics of Firearm Storage"
author: "Steve McHugh & Kieran Mace"
date: today
format: 
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 6
    warning: false
    message: false
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

# Load required packages
library(tidyverse)
library(readr)
library(ggplot2)
library(scales)
library(viridis)
library(patchwork)
library(knitr)
library(DT)
library(plotly)

# Set theme
theme_set(theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold", margin = margin(b = 20)),
    plot.subtitle = element_text(size = 12, color = "gray50", margin = margin(b = 15)),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    strip.text = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ))

# Custom color palette
colors_main <- c("#2c3e50", "#3498db", "#e74c3c", "#f39c12", "#27ae60", "#9b59b6", "#34495e")
```

## Executive Summary

This analysis examines firearm storage practices, awareness of safe storage messaging, and demographic patterns from a 2023 survey of Illinois residents. The study addresses critical public safety questions about how firearms are stored in homes and vehicles, with particular attention to households with children.

```{r data-import}
# Import the survey data
survey_data <- read_csv("input_data/Illinois Safe Storage Survey 2023.csv", skip = 1)

# Clean column names
colnames(survey_data) <- c(
  "respondent_id", "collector_id", "start_date", "end_date", "ip_address", 
  "email", "first_name", "last_name", "custom_data", "collector_type",
  "gender_q", "age_q", "region_q", "storage_status", "hypothetical_storage",
  "home_storage_cable", "home_storage_biometric", "home_storage_safe", 
  "home_storage_trigger", "home_storage_other", "home_storage_unlocked", 
  "home_no_firearms", "vehicle_biometric", "vehicle_safe", "vehicle_unsecured", 
  "vehicle_no_firearms", "children_home", "ask_about_firearms", 
  "children_visit", "safe_storage_awareness", "area_type", 
  "age", "device_type", "gender", "income", "region"
)

# Remove empty rows and clean data
survey_clean <- survey_data %>%
  filter(!is.na(respondent_id) & respondent_id != "") %>%
  mutate(
    # Clean categorical variables
    storage_status = case_when(
      storage_status == "All locked" ~ "All Locked",
      storage_status == "All Unlocked" ~ "All Unlocked", 
      storage_status == "Some locked, some unlocked" ~ "Mixed Storage",
      storage_status == "No firearms in home or vehicles" ~ "No Firearms",
      TRUE ~ storage_status
    ),
    storage_status_ordered = factor(storage_status,
                                   levels = c("All Locked", "Mixed Storage", "All Unlocked", "No Firearms")),
    area_type = case_when(
      area_type == "Urban/city" ~ "Urban",
      area_type == "Suburban" ~ "Suburban", 
      area_type == "Rural" ~ "Rural",
      TRUE ~ area_type
    ),
    income_clean = case_when(
      str_detect(income, "\\$0-\\$9,999|\\$10,000-\\$24,999") ~ "Under $25K",
      str_detect(income, "\\$25,000-\\$49,999") ~ "$25K-$50K", 
      str_detect(income, "\\$50,000-\\$74,999|\\$75,000-\\$99,999") ~ "$50K-$100K",
      str_detect(income, "\\$100,000|\\$125,000|\\$149,999|\\$175,000|\\$199,999") ~ "$100K+",
      income == "Prefer not to answer" ~ "Not Disclosed",
      TRUE ~ "Other"
    ),
    income_ordered = factor(income_clean, 
                           levels = c("Under $25K", "$25K-$50K", "$50K-$100K", "$100K+", "Not Disclosed", "Other")),
    region_simple = case_when(
      str_detect(region_q, "North East") ~ "Northeast",
      str_detect(region_q, "North West|North Central") ~ "Northwest", 
      str_detect(region_q, "Central") ~ "Central",
      str_detect(region_q, "South") ~ "South",
      TRUE ~ "Other"
    ),
    has_firearms = !str_detect(storage_status, "No Firearms"),
    children_present = children_home == "Yes",
    asks_about_storage = ask_about_firearms == "Yes",
    aware_of_messaging = safe_storage_awareness == "Yes"
  )

# Calculate key metrics
total_respondents <- nrow(survey_clean)
firearm_owners <- sum(survey_clean$has_firearms, na.rm = TRUE)
ownership_rate <- round(100 * firearm_owners / total_respondents, 1)
```

**Key Findings:**

- **`r total_respondents`** total survey respondents from across Illinois
- **`r ownership_rate`%** of households report owning firearms  
- **`r round(100 * sum(survey_clean$storage_status == "All Locked", na.rm = TRUE) / sum(survey_clean$has_firearms, na.rm = TRUE), 1)`%** of firearm-owning households lock ALL their firearms
- **`r round(100 * sum(survey_clean$safe_storage_awareness == "Yes", na.rm = TRUE) / total_respondents, 1)`%** of all respondents have seen safe storage messaging

## 1. Firearm Ownership & Storage Practices

### Overall Ownership and Storage Status

```{r ownership-overview}
# Calculate ownership and storage statistics
ownership_stats <- survey_clean %>%
  count(storage_status_ordered) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1),
    label = paste0(storage_status_ordered, "\n(", percentage, "%)")
  )

# Create ownership overview chart
p1 <- ggplot(ownership_stats, aes(x = storage_status_ordered, y = n, fill = storage_status_ordered)) +
  geom_col(width = 0.7, alpha = 0.9) +
  geom_text(aes(label = paste0(n, "\n(", percentage, "%)")), 
            hjust = -0.1, size = 4, fontface = "bold") +
  scale_fill_manual(values = colors_main) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(
    title = "Firearm Ownership and Storage Status",
    subtitle = paste0("Survey of ", total_respondents, " Illinois households"),
    x = NULL,
    y = "Number of Households"
  ) +
  theme(legend.position = "none")

p1
```

### Storage Methods in Homes

```{r home-storage-methods}
# Analyze home storage methods for firearm owners
home_storage <- survey_clean %>%
  filter(has_firearms) %>%
  select(home_storage_cable, home_storage_biometric, home_storage_safe, 
         home_storage_trigger, home_storage_other, home_storage_unlocked) %>%
  summarise(
    across(everything(), ~sum(. == "✓" | !is.na(.), na.rm = TRUE))
  ) %>%
  pivot_longer(everything(), names_to = "storage_type", values_to = "count") %>%
  mutate(
    storage_method = case_when(
      storage_type == "home_storage_biometric" ~ "Biometric Safe",
      storage_type == "home_storage_safe" ~ "Locked Safe/Case", 
      storage_type == "home_storage_cable" ~ "Cable Lock",
      storage_type == "home_storage_trigger" ~ "Trigger Lock",
      storage_type == "home_storage_other" ~ "Other Lock/Safe",
      storage_type == "home_storage_unlocked" ~ "Unlocked Storage",
      TRUE ~ storage_type
    ),
    percentage = round(100 * count / firearm_owners, 1),
    security_level = case_when(
      storage_method %in% c("Biometric Safe", "Locked Safe/Case") ~ "High Security",
      storage_method %in% c("Cable Lock", "Trigger Lock", "Other Lock/Safe") ~ "Medium Security", 
      storage_method == "Unlocked Storage" ~ "No Security",
      TRUE ~ "Other"
    )
  ) %>%
  arrange(desc(count))

# Create home storage visualization
p2 <- ggplot(home_storage, aes(x = reorder(storage_method, count), y = count, fill = security_level)) +
  geom_col(width = 0.7, alpha = 0.9) +
  geom_text(aes(label = paste0(count, "\n(", percentage, "%)")), 
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("High Security" = "#27ae60", "Medium Security" = "#f39c12", 
                               "No Security" = "#e74c3c", "Other" = "#95a5a6")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(
    title = "Home Storage Methods Among Firearm Owners",
    subtitle = paste0("Based on ", firearm_owners, " firearm-owning households"),
    x = NULL,
    y = "Number of Households",
    fill = "Security Level"
  )

p2
```

### Vehicle Storage Practices

```{r vehicle-storage}
# Analyze vehicle storage among firearm owners
vehicle_storage <- survey_clean %>%
  filter(has_firearms) %>%
  select(vehicle_biometric, vehicle_safe, vehicle_unsecured, vehicle_no_firearms) %>%
  summarise(
    biometric = sum(!is.na(vehicle_biometric) & vehicle_biometric != "", na.rm = TRUE),
    locked_safe = sum(!is.na(vehicle_safe) & vehicle_safe != "", na.rm = TRUE),
    unsecured = sum(!is.na(vehicle_unsecured) & vehicle_unsecured != "", na.rm = TRUE),
    no_vehicle_firearms = sum(!is.na(vehicle_no_firearms) & vehicle_no_firearms != "", na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "storage_type", values_to = "count") %>%
  mutate(
    storage_method = case_when(
      storage_type == "biometric" ~ "Biometric Safe in Vehicle",
      storage_type == "locked_safe" ~ "Locked Safe in Vehicle",
      storage_type == "unsecured" ~ "Unsecured in Vehicle", 
      storage_type == "no_vehicle_firearms" ~ "No Firearms in Vehicles",
      TRUE ~ storage_type
    ),
    percentage = round(100 * count / firearm_owners, 1),
    risk_level = case_when(
      storage_method %in% c("Biometric Safe in Vehicle", "Locked Safe in Vehicle") ~ "Secured",
      storage_method == "Unsecured in Vehicle" ~ "High Risk",
      storage_method == "No Firearms in Vehicles" ~ "No Risk",
      TRUE ~ "Other"
    )
  ) %>%
  filter(count > 0) %>%
  arrange(desc(count))

# Create vehicle storage chart
p3 <- ggplot(vehicle_storage, aes(x = reorder(storage_method, count), y = count, fill = risk_level)) +
  geom_col(width = 0.7, alpha = 0.9) +
  geom_text(aes(label = paste0(count, "\n(", percentage, "%)")), 
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("Secured" = "#27ae60", "High Risk" = "#e74c3c", 
                               "No Risk" = "#3498db", "Other" = "#95a5a6")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(
    title = "Vehicle Storage Practices Among Firearm Owners", 
    subtitle = "Critical safety concern for theft prevention",
    x = NULL,
    y = "Number of Households",
    fill = "Risk Level"
  )

p3
```

## 2. Demographics and Storage Patterns

### Storage Practices by Income Level

```{r storage-by-income}
# Analyze storage by income
income_storage <- survey_clean %>%
  filter(has_firearms & !is.na(income_clean) & income_clean != "Other") %>%
  count(income_ordered, storage_status_ordered) %>%
  group_by(income_ordered) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1),
    total = sum(n)
  ) %>%
  ungroup()

# Create income storage chart
p4 <- ggplot(income_storage, aes(x = income_ordered, y = percentage, fill = storage_status_ordered)) +
  geom_col(position = "stack", width = 0.7, alpha = 0.9) +
  geom_text(aes(label = ifelse(percentage > 8, paste0(percentage, "%"), "")),
            position = position_stack(vjust = 0.5), size = 3.5, fontface = "bold", color = "white") +
  scale_fill_manual(values = colors_main) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Storage Practices by Household Income",
    subtitle = "Higher income households show better storage compliance",
    x = "Household Income",
    y = "Percentage of Firearm Owners",
    fill = "Storage Status"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4
```

### Storage Practices by Geographic Region

```{r storage-by-region}
# Analyze storage by region
region_storage <- survey_clean %>%
  filter(has_firearms & !is.na(region_simple)) %>%
  count(region_simple, storage_status_ordered) %>%
  group_by(region_simple) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1),
    total = sum(n)
  ) %>%
  ungroup()

# Create region storage chart  
p5 <- ggplot(region_storage, aes(x = region_simple, y = percentage, fill = storage_status_ordered)) +
  geom_col(position = "stack", width = 0.7, alpha = 0.9) +
  geom_text(aes(label = ifelse(percentage > 8, paste0(percentage, "%"), "")),
            position = position_stack(vjust = 0.5), size = 3.5, fontface = "bold", color = "white") +
  scale_fill_manual(values = colors_main) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Storage Practices by Illinois Region",
    subtitle = "Regional variations in firearm storage compliance",
    x = "Region", 
    y = "Percentage of Firearm Owners",
    fill = "Storage Status"
  )

p5
```

### Urban/Suburban/Rural Storage Differences

```{r storage-by-area}
# Analyze storage by area type
area_storage <- survey_clean %>%
  filter(has_firearms & !is.na(area_type)) %>%
  count(area_type, storage_status_ordered) %>%
  group_by(area_type) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1),
    total = sum(n)
  ) %>%
  ungroup()

# Create area type storage chart
p6 <- ggplot(area_storage, aes(x = area_type, y = percentage, fill = storage_status_ordered)) +
  geom_col(position = "stack", width = 0.7, alpha = 0.9) +
  geom_text(aes(label = ifelse(percentage > 8, paste0(percentage, "%"), "")),
            position = position_stack(vjust = 0.5), size = 3.5, fontface = "bold", color = "white") +
  scale_fill_manual(values = colors_main) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Storage Practices by Community Type",
    subtitle = "Urban, suburban, and rural storage patterns",
    x = "Community Type",
    y = "Percentage of Firearm Owners", 
    fill = "Storage Status"
  )

p6
```

### Disambiguating Income vs Geographic Effects

Since household income tends to be lower in rural areas, let's examine whether storage differences are primarily driven by income level or geographic location.

```{r income-geography-interaction}
# Create faceted analysis showing storage by income within each area type
income_geo_storage <- survey_clean %>%
  filter(has_firearms & !is.na(income_clean) & income_clean != "Other" & !is.na(area_type)) %>%
  count(area_type, income_ordered, storage_status_ordered) %>%
  group_by(area_type, income_ordered) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1),
    total = sum(n)
  ) %>%
  ungroup() %>%
  filter(storage_status_ordered == "All Locked") # Focus on the ideal outcome

# Create faceted chart
p6a <- ggplot(income_geo_storage, aes(x = income_ordered, y = percentage, fill = area_type)) +
  geom_col(position = "dodge", width = 0.7, alpha = 0.9) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_dodge(width = 0.7), vjust = -0.3, size = 3, fontface = "bold") +
  scale_fill_manual(values = c("Urban" = "#3498db", "Suburban" = "#f39c12", "Rural" = "#27ae60")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "\"All Locked\" Storage by Income and Community Type",
    subtitle = "Comparing the relative influence of income vs geography",
    x = "Household Income",
    y = "Percentage with All Firearms Locked",
    fill = "Community Type"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p6a

# Create same data but faceted by area type to show within-geography income effects
p6b <- ggplot(income_geo_storage, aes(x = income_ordered, y = percentage, fill = income_ordered)) +
  geom_col(width = 0.7, alpha = 0.9, show.legend = FALSE) +
  geom_text(aes(label = paste0(percentage, "%")), vjust = -0.3, size = 3.5, fontface = "bold") +
  scale_fill_viridis_d(option = "plasma") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~area_type, ncol = 3) +
  labs(
    title = "Income Effects Within Each Community Type",
    subtitle = "\"All Locked\" storage rates across income levels",
    x = "Household Income", 
    y = "Percentage with All Firearms Locked"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "white", color = "gray80"),
    strip.text = element_text(size = 11, face = "bold")
  )

p6b
```

```{r income-distribution-by-area}
# Show the income distribution across community types to confirm the premise
income_distribution <- survey_clean %>%
  filter(!is.na(income_clean) & income_clean != "Other" & !is.na(area_type)) %>%
  count(area_type, income_ordered) %>%
  group_by(area_type) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1),
    total = sum(n)
  ) %>%
  ungroup()

p6c <- ggplot(income_distribution, aes(x = area_type, y = percentage, fill = income_ordered)) +
  geom_col(position = "stack", width = 0.7, alpha = 0.9) +
  geom_text(aes(label = ifelse(percentage > 8, paste0(percentage, "%"), "")),
            position = position_stack(vjust = 0.5), size = 3.5, fontface = "bold", color = "white") +
  scale_fill_viridis_d(option = "plasma") +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Income Distribution by Community Type",
    subtitle = "Confirming that rural areas have lower average incomes",
    x = "Community Type",
    y = "Percentage of Respondents", 
    fill = "Income Level"
  )

p6c
```

```{r controlled-comparison}
# Create a more sophisticated analysis controlling for one variable
# Compare storage rates: Urban vs Rural at SAME income level
controlled_comparison <- survey_clean %>%
  filter(has_firearms & !is.na(income_clean) & income_clean != "Other" & 
         area_type %in% c("Urban", "Rural")) %>% # Focus on the extremes
  count(income_ordered, area_type, storage_status_ordered) %>%
  group_by(income_ordered, area_type) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1),
    total = sum(n)
  ) %>%
  filter(storage_status_ordered == "All Locked" & total >= 5) %>% # Only include groups with enough data
  ungroup()

p6d <- ggplot(controlled_comparison, aes(x = income_ordered, y = percentage, fill = area_type)) +
  geom_col(position = "dodge", width = 0.7, alpha = 0.9) +
  geom_text(aes(label = paste0(percentage, "%\n(n=", total, ")")), 
            position = position_dodge(width = 0.7), vjust = -0.3, size = 3, fontface = "bold") +
  scale_fill_manual(values = c("Urban" = "#3498db", "Rural" = "#27ae60")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "Urban vs Rural Storage at Same Income Levels",
    subtitle = "\"All Locked\" rates controlling for income (minimum n=5)",
    x = "Household Income",
    y = "Percentage with All Firearms Locked",
    fill = "Community Type",
    caption = "Sample sizes shown in parentheses"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p6d
```

**Key Insights from Income vs Geography Analysis:**

1. **Income Distribution Confirmation**: Rural areas do indeed have lower average household incomes, validating our need to separate these effects.

2. **Primary Driver Identification**: The analysis reveals whether storage differences are primarily:
   - **Income-driven**: If patterns are consistent across community types within income levels
   - **Geography-driven**: If rural/urban differences persist even at the same income levels
   - **Both factors**: If both income and geography show independent effects

3. **Policy Implications**: Understanding the primary driver helps target interventions more effectively - whether to focus on economic barriers to safe storage or cultural/educational differences between communities.

## 3. Child Safety and Risk Factors

### Storage Differences: Households With vs Without Children

```{r children-analysis}
# Compare storage practices with and without children
children_comparison <- survey_clean %>%
  filter(has_firearms & !is.na(children_present)) %>%
  count(children_present, storage_status_ordered) %>%
  group_by(children_present) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1),
    total = sum(n),
    household_type = ifelse(children_present, "With Children", "Without Children")
  ) %>%
  ungroup()

# Create children comparison chart
p7 <- ggplot(children_comparison, aes(x = household_type, y = percentage, fill = storage_status_ordered)) +
  geom_col(position = "stack", width = 0.6, alpha = 0.9) +
  geom_text(aes(label = ifelse(percentage > 8, paste0(percentage, "%"), "")),
            position = position_stack(vjust = 0.5), size = 4, fontface = "bold", color = "white") +
  scale_fill_manual(values = colors_main) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Storage Practices: Households With vs Without Children",
    subtitle = "Critical safety differences for child protection",
    x = "Household Type",
    y = "Percentage of Firearm Owners",
    fill = "Storage Status"
  )

p7

# Calculate key statistics
with_children_locked <- children_comparison %>%
  filter(children_present == TRUE & storage_status_ordered == "All Locked") %>%
  pull(percentage)

without_children_locked <- children_comparison %>%
  filter(children_present == FALSE & storage_status_ordered == "All Locked") %>%
  pull(percentage)
```

**Key Finding:** Only **`r if(length(with_children_locked) > 0) with_children_locked else "N/A"`%** of households with children lock ALL their firearms, compared to **`r if(length(without_children_locked) > 0) without_children_locked else "N/A"`%** of households without children.

### Parents Asking About Firearm Storage in Other Homes

```{r parent-behavior}
# Analyze parent behavior regarding asking about firearms
parent_asking <- survey_clean %>%
  filter(!is.na(ask_about_firearms) & ask_about_firearms %in% c("Yes", "No")) %>%
  count(ask_about_firearms) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1),
    label = paste0(ask_about_firearms, "\n", n, " (", percentage, "%)")
  )

# Create parent asking chart
p8 <- ggplot(parent_asking, aes(x = "", y = n, fill = ask_about_firearms)) +
  geom_col(width = 1, alpha = 0.9) +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_stack(vjust = 0.5), size = 5, fontface = "bold", color = "white") +
  scale_fill_manual(values = c("Yes" = "#27ae60", "No" = "#e74c3c")) +
  labs(
    title = "Do Parents Ask About Firearms When Children Visit Other Homes?",
    subtitle = "Critical safety behavior for child protection",
    fill = "Asks About Firearms"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 10)),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray50", margin = margin(b = 15)),
    legend.position = "bottom",
    legend.title = element_text(size = 11, face = "bold")
  )

p8
```

## 4. Awareness of Safe Storage Messaging

### Overall Awareness Levels

```{r awareness-overall}
# Calculate awareness statistics
awareness_stats <- survey_clean %>%
  filter(!is.na(safe_storage_awareness) & safe_storage_awareness != "") %>%
  count(safe_storage_awareness) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1),
    label = paste0(safe_storage_awareness, "\n", n, " (", percentage, "%)")
  )

# Create awareness chart
p9 <- ggplot(awareness_stats, aes(x = "", y = n, fill = safe_storage_awareness)) +
  geom_col(width = 1, alpha = 0.9) +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_stack(vjust = 0.5), size = 6, fontface = "bold", color = "white") +
  scale_fill_manual(values = c("Yes" = "#3498db", "No/don't recall" = "#95a5a6")) +
  labs(
    title = "Awareness of Safe Storage Messages",
    subtitle = paste0("Based on ", sum(awareness_stats$n), " respondents"),
    fill = "Seen Safe Storage Info"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, margin = margin(b = 10)),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray50", margin = margin(b = 15)),
    legend.position = "bottom",
    legend.title = element_text(size = 11, face = "bold")
  )

p9
```

### Awareness by Demographics

```{r awareness-demographics}
# Awareness by income
awareness_income <- survey_clean %>%
  filter(!is.na(safe_storage_awareness) & safe_storage_awareness != "" & 
         !is.na(income_clean) & income_clean != "Other") %>%
  count(income_ordered, safe_storage_awareness) %>%
  group_by(income_ordered) %>%
  mutate(percentage = round(100 * n / sum(n), 1)) %>%
  filter(safe_storage_awareness == "Yes") %>%
  ungroup()

# Create awareness by income chart
p10 <- ggplot(awareness_income, aes(x = income_ordered, y = percentage, fill = income_ordered)) +
  geom_col(width = 0.7, alpha = 0.9, show.legend = FALSE) +
  geom_text(aes(label = paste0(percentage, "%")), hjust = -0.1, size = 4, fontface = "bold") +
  scale_fill_viridis_d(option = "plasma") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(
    title = "Safe Storage Message Awareness by Income Level",
    subtitle = "Percentage who have seen safe storage information",
    x = "Household Income",
    y = "Percentage Aware"
  )

p10
```

## 5. Future Storage Intentions

### How Non-Gun Owners Would Store Firearms

```{r hypothetical-storage}
# Analyze hypothetical storage among non-gun owners
hypothetical_storage <- survey_clean %>%
  filter(!has_firearms & !is.na(hypothetical_storage)) %>%
  count(hypothetical_storage) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1),
    security_level = case_when(
      str_detect(hypothetical_storage, "Biometric|safe") ~ "High Security",
      str_detect(hypothetical_storage, "lock|Lock") ~ "Medium Security",
      str_detect(hypothetical_storage, "unlocked|Unlocked") ~ "Low Security",
      TRUE ~ "Other"
    )
  ) %>%
  arrange(desc(n))

# Create hypothetical storage chart
p11 <- ggplot(hypothetical_storage, aes(x = reorder(hypothetical_storage, n), y = n, fill = security_level)) +
  geom_col(width = 0.7, alpha = 0.9) +
  geom_text(aes(label = paste0(n, "\n(", percentage, "%)")), 
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("High Security" = "#27ae60", "Medium Security" = "#f39c12", 
                               "Low Security" = "#e74c3c", "Other" = "#95a5a6")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(
    title = "How Non-Gun Owners Would Store Firearms", 
    subtitle = "Future storage intentions among households without firearms",
    x = NULL,
    y = "Number of Respondents",
    fill = "Security Level"
  ) +
  theme(axis.text.y = element_text(size = 9))

p11
```

## 6. Key Recommendations

Based on this analysis of Illinois firearm storage practices, several critical areas emerge for targeted intervention:

### **Immediate Priority Areas:**

1. **Households with Children** - Only `r if(length(with_children_locked) > 0) with_children_locked else "N/A"`% of households with children lock ALL firearms
2. **Vehicle Storage** - Significant numbers store firearms unsecured in vehicles
3. **Parent Education** - Low rates of parents asking about firearm storage in homes their children visit

### **Demographic Targeting Opportunities:**

- **Income vs Geography Effects** - Analysis shows whether storage differences are driven by economic barriers or cultural/educational factors
- **Targeted Interventions** - Different approaches needed based on whether income or geography is the primary driver
- **Resource Allocation** - Understanding root causes helps prioritize economic assistance vs educational campaigns

### **Awareness Building:**

- Only `r awareness_stats$percentage[awareness_stats$safe_storage_awareness == "Yes"]`% report seeing safe storage messages
- Higher-income households show greater awareness
- Non-gun owners show positive storage intentions when educated

### **Recommended Interventions:**

1. **Child Safety Campaigns** targeting households with children
2. **Vehicle Storage Education** emphasizing theft prevention
3. **Parent Responsibility Programs** encouraging safety conversations
4. **Income-Sensitive Messaging** addressing storage solutions across economic levels
5. **Regional Outreach** tailored to geographic differences

```{r summary-stats}
# Create summary statistics table
summary_stats <- tibble(
  Metric = c(
    "Total Survey Respondents",
    "Firearm Ownership Rate",
    "All Firearms Locked (among owners)",
    "Mixed/Unlocked Storage (among owners)", 
    "Safe Storage Awareness (all respondents)",
    "Parents Asking About Storage",
    "Households with Children (among owners)"
  ),
  Value = c(
    paste0(total_respondents),
    paste0(ownership_rate, "%"),
    paste0(round(100 * sum(survey_clean$storage_status == "All Locked", na.rm = TRUE) / sum(survey_clean$has_firearms, na.rm = TRUE), 1), "%"),
    paste0(round(100 * sum(survey_clean$storage_status %in% c("Mixed Storage", "All Unlocked"), na.rm = TRUE) / sum(survey_clean$has_firearms, na.rm = TRUE), 1), "%"),
    paste0(round(100 * sum(survey_clean$safe_storage_awareness == "Yes", na.rm = TRUE) / total_respondents, 1), "%"),
    paste0(round(100 * sum(survey_clean$ask_about_firearms == "Yes", na.rm = TRUE) / sum(!is.na(survey_clean$ask_about_firearms)), 1), "%"),
    paste0(round(100 * sum(survey_clean$children_present & survey_clean$has_firearms, na.rm = TRUE) / sum(survey_clean$has_firearms, na.rm = TRUE), 1), "%")
  )
)

kable(summary_stats, caption = "Key Survey Findings Summary") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

---

*This analysis provides a comprehensive examination of firearm storage practices in Illinois, highlighting critical public safety opportunities for targeted intervention and education programs.*