---
title: Determining How long it will take to get your EAD from USCIS
author: Kieran Mace
date: '2018-02-13'
categories:
  - R
slug: determining-how-long-it-will-take-to-get-your-ead-from-uscis
output:
  html_document:
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Scraping data off the USCIS website

I found some code online (https://github.com/co89757/USCISCasePoll/blob/master/poll_uscis.py) to scrape the USCIS website for status updates. I used it to collect the case status for every 10th case between last October and today.

```{python, eval=FALSE}

from pyquery import PyQuery as pq
import requests
import smtplib
import os
import sys
import os.path
import re
import pandas as pd
import feather

STATUS_OK = 0
STATUS_ERROR = -1
FILENAME_LASTSTATUS = os.path.join(sys.path[0], "LAST_STATUS_{0}.txt")
mynum = 1890048782 # THis is my case number

def poll_optstatus(casenumber):
    """
    poll USCIS case status given receipt number (casenumber)
    Args:
        param1: casenumber the case receipt number
    Returns:
        a tuple (status, details) containing status and detailed info
    Raise:
        error:
    """
    headers = {
        'Accept': 'text/html, application/xhtml+xml, image/jxr, */*',
        'Accept-Encoding': 'gzip, deflate',
        'Accept-Language':
        'en-US, en; q=0.8, zh-Hans-CN; q=0.5, zh-Hans; q=0.3',
        'Cache-Control': 'no-cache',
        'Connection': 'Keep-Alive',
        'Content-Type': 'application/x-www-form-urlencoded',
        'Host': 'egov.uscis.gov',
        'Referer': 'https://egov.uscis.gov/casestatus/mycasestatus.do',
        'User-Agent':
        'Mozilla/5.0 (Windows NT 10.0; Win64; x64) AppleWebKit/537.36 (KHTML, like Gecko) Chrome/46.0.2486.0 Safari/537.36 Edge/13.10586'
    }
    url = "https://egov.uscis.gov/casestatus/mycasestatus.do"
    data = {"appReceiptNum": casenumber, 'caseStatusSearchBtn': 'CHECK+STATUS'}

    res = requests.post(url, data=data, headers=headers)
    doc = pq(res.text)
    status = doc('h1').text()
    code = STATUS_OK if status else STATUS_ERROR
    details = doc('.text-center p').text()
    return (code, status, details)

# Get every 10th case status
case_nums = ['YSC' + str(i) for i in range(1890038932, 1890079632)]
vals = [poll_optstatus(case) for case in case_nums]

df = pd.DataFrame.from_records(vals)
df['case'] = case_nums

feather.write_dataframe(df, "uscis.feather")
```


## Data Transformations and cleaning

```{r, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(feather)
library(lubridate)
library(magrittr)
```


```{r, warning=FALSE}
cases = read_feather("uscis.feather")
colnames(cases)[1:3] = c('code', 'status', 'details')

my_case_numeric = 1890048782
my_case_date = parse_date("2017-11-21")
```



I filtered out all cases that were rejected or cancelled for any reason. 

This leaves behind cases that have either been __processed__, or cases that are still  __unprocessed__.

```{r}
cases %<>% 
  filter(code == 0 & ((status == "Case Was Received" & grepl("765", details)) | 
         status == 'New Card Is Being Produced')) %>%
  mutate(status_date = parse_date(word(details,2,4),format="%B %d, %Y,"),
         case_numeric = as.numeric(str_extract(case,"[0-9]+")))
  
```


## Unprocessed cases

First lets look at the distribution of cases that have not yet been processed, including mine:

```{r pressure, echo=FALSE}
  cases %>%
  filter(status == "Case Was Received") -> cases_pending
cases_pending %>%
  ggplot(aes(x=status_date)) + 
  geom_density(fill = '#F8766D') + 
  geom_vline(xintercept=as.numeric(my_case_date)) + 
  ggtitle('Distribution of unprocessed cases', 'My case is the verticle black line') +
  xlab('Date application recieved') +
  theme_classic()
```

We can see that UCSIS is falling behind on their promise to process applications between 75-90 days. they seem to have finished cases that were submitted in the beginning of November, which was 106 days ago. 

## Newly processed cases

Now we will look at the cases that USCIS has recently completed.

```{r}

fit <- smooth.spline(cases_pending$case_numeric, cases_pending$status_date)
cases$predicted_receival_date = as.Date(
                                  predict(fit, cases$case_numeric)$y, 
                                  origin="1970-01-01", 
                                  tz="EST")

cases_pending$predicted_receival_date = as.Date(
                                          predict(fit, cases_pending$case_numeric)$y, 
                                          origin="1970-01-01", 
                                          tz="EST")

ggplot(cases, aes(x=predicted_receival_date, 
                  y = status, 
                  color = status)) + 
  geom_jitter() + 
  geom_vline(xintercept = my_case_date) +
  ggtitle('Case Status by Date Recieved', 'My case is indicated by the vertical line') +
  xlab('Date Recieved') +
  ylab('Case Status') +
  theme_classic() +
  scale_color_discrete(name="Process Status",
                         breaks=c("Case Was Received", "New Card Is Being Produced"),
                         labels=c("Unprocessed", "Processed"))
```

From this plot, it seems like cases around mid November are currently being processed, my official date is November 21st. 

## Conclusion

From this data, I can make the following observations:

* Cases with the same date as mine have just started being processed
* It seems like it takes about 21 days for an application date to go from __completely unprocessed__ to __completely processed__.

__**Therefore:**__

* There is a small, but non-zero chance it will be processed in the next 4 days. 
* There is a 50% chance it will be processed in the next 10 days. 
* There is close to a 100% chance that my case will be processed in the next 21 days. 


## Appendix

### Converting Case number to date recieved

Cases with "New Card Is Being Produced" catagory do not indicate the date that those cases were first recieved, therefore I needed to convert the `case number` to the date the application was recieved. This is well approximated using a simple spline:

```{r}
ggplot(cases_pending, aes(x=case_numeric, y = status_date)) + geom_point() +
  geom_line(aes(y=predicted_receival_date), color = 'red') + xlab('Case Number') + ylab('Case Recieved') + theme_classic()
```